---
title: "Analysis"
output: html_notebook
---

# Load Models

```{r packages, include=FALSE}

library(stringr)
library(dplyr)
library(ergm)
library(xtable)
BASE_PATH <- "~/math5005/msresearch/outputs/subphase10/"
```


```{r functions}

# for some reason the network stats aren't coming through
load_network_rdata <- function(fname) {
  print(to_load <- paste0(BASE_PATH, fname, ".RData"))
  l_network_stats <- get(load(to_load))
  network <- ergm_output # ergm_output only loaded if the rest is loaded
  return(network)
}

load_stats_rdata <- function(fname1, fname2, fname3) {
  all_files <- list()
  for (fname in c(fname1, fname2, fname3)) {
    print(to_load <- paste0(BASE_PATH, fname, ".RData"))
    all_files[fname] <- get(load(to_load))
  }
  return(all_files)
}

# NOTE: this depends on order of stats being read is zeros -> zeros_edges -> MPLE
tabulate_runtimes <- function(df, stats) {
  tmp_df <- rbind(stats[1][[1]]$timings, stats[2][[1]]$timings, stats[3][[1]]$timings)
  row.names(tmp_df) <- names(stats)
  df <- rbind(df, tmp_df)
}

tabulate_parameters <- function(df, stats) {
  tmp_df <- rbind(stats[1][[1]]$coef, stats[2][[1]]$coef, stats[3][[1]]$coef)
  row.names(tmp_df) <- names(stats)
  df <- rbind(df, tmp_df)
}

```

```{r load-ecoli-models, include=FALSE}

# stochastic approximation
ecoli_sa_zeros <- load_network_rdata("ecoli2.Stochastic-Approximation.zeros")
ecoli_sa_zero_edges <- load_network_rdata("ecoli2.Stochastic-Approximation.zeros_and_edges")
ecoli_sa_mple <- load_network_rdata("ecoli2.Stochastic-Approximation.MPLE")
stats_ecoli_sa <- load_stats_rdata("ecoli2.Stochastic-Approximation.zeros", "ecoli2.Stochastic-Approximation.zeros_and_edges", "ecoli2.Stochastic-Approximation.MPLE")
# equilibrium expectation
ecoli_ee_zeros <- load_network_rdata("ecoli2.EE.zeros")
ecoli_ee_zero_edges <- load_network_rdata("ecoli2.EE.zeros_and_edges")
ecoli_ee_mple <- load_network_rdata("ecoli2.EE.MPLE")
stats_ecoli_ee <- load_stats_rdata("ecoli2.EE.zeros", "ecoli2.EE.zeros_and_edges", "ecoli2.EE.MPLE")
# MCMLE
ecoli_mcmle_zeros <- load_network_rdata("ecoli2.MCMLE.zeros")
ecoli_mcmle_zero_edges <- load_network_rdata("ecoli2.MCMLE.zeros_and_edges")
ecoli_mcmle_mple <- load_network_rdata("ecoli2.MCMLE.MPLE")
stats_ecoli_mcmle <- load_stats_rdata("ecoli2.MCMLE.zeros", "ecoli2.MCMLE.zeros_and_edges", "ecoli2.MCMLE.MPLE")

```

```{r load-ecoli-selfstudy-models, include=FALSE}

# stochastic approximation
self_ecoli_sa_zeros <- load_network_rdata("ecoli2.self.study.Stochastic-Approximation.zeros")
self_ecoli_sa_zero_edges <- load_network_rdata("ecoli2.self.study.Stochastic-Approximation.zeros_and_edges")
self_ecoli_sa_mple <- load_network_rdata("ecoli2.self.study.Stochastic-Approximation.MPLE")
self_stats_ecoli_sa <- load_stats_rdata("ecoli2.self.study.Stochastic-Approximation.zeros", "ecoli2.self.study.Stochastic-Approximation.zeros_and_edges", "ecoli2.self.study.Stochastic-Approximation.MPLE")
# equilibrium expectation
self_ecoli_ee_zeros <- load_network_rdata("ecoli2.self.study.EE.zeros")
self_ecoli_ee_zero_edges <- load_network_rdata("ecoli2.self.study.EE.zeros_and_edges")
self_ecoli_ee_mple <- load_network_rdata("ecoli2.self.study.EE.MPLE")
self_stats_ecoli_ee <- load_stats_rdata("ecoli2.self.study.EE.zeros", "ecoli2.self.study.EE.zeros_and_edges", "ecoli2.self.study.EE.MPLE")
# MCMLE
self_ecoli_mcmle_zeros <- load_network_rdata("ecoli2.self.study.MCMLE.zeros")
self_ecoli_mcmle_zero_edges <- load_network_rdata("ecoli2.self.study.MCMLE.zeros_and_edges")
self_ecoli_mcmle_mple <- load_network_rdata("ecoli2.self.study.MCMLE.MPLE")
self_stats_ecoli_mcmle <- load_stats_rdata("ecoli2.self.study.MCMLE.zeros", "ecoli2.self.study.MCMLE.zeros_and_edges", "ecoli2.self.study.MCMLE.MPLE")

```

Kapferer models have no MCMLE method right now since it doesn;t seem to be working.

```{r kapferer1-models, eval=FALSE, include=FALSE}

# stochastic approximation
kap1_sa_zeros <- load_network_rdata("kapferer.Stochastic-Approximation.zeros")
kap1_sa_zero_edges <- load_network_rdata("kapferer.Stochastic-Approximation.zeros_and_edges")
kap1_sa_mple <- load_network_rdata("kapferer.Stochastic-Approximation.MPLE")
stats_kap1_sa <- load_stats_rdata("kapferer.Stochastic-Approximation.zeros", "kapferer.Stochastic-Approximation.zeros_and_edges", "kapferer.Stochastic-Approximation.MPLE")
# equilibrium expectation
kap1_ee_zeros <- load_network_rdata("kapferer.EE.zeros")
kap1_ee_zero_edges <- load_network_rdata("kapferer.EE.zeros_and_edges")
kap1_ee_mple <- load_network_rdata("kapferer.EE.MPLE")
stats_kap1_ee <- load_stats_rdata("kapferer.EE.zeros", "kapferer.EE.zeros_and_edges", "kapferer.EE.MPLE")
# MCMLE
# kap1_mcmle_zeros <- load_network_rdata("kapferer.MCMLE.zeros")
# kap1_mcmle_zero_edges <- load_network_rdata("kapferer.MCMLE.zeros_and_edges")
# kap1_mcmle_mple <- load_network_rdata("kapferer.MCMLE.MPLE")
# stats_kap1_mcmle <- load_stats_rdata("kapferer.Stochastic-Approximation.zeros", "kapferer.Stochastic-Approximation.zeros_and_edges", "kapferer.Stochastic-Approximation.MPLE")

```

```{r}
# stochastic approximation
kap2_sa_zeros <- load_network_rdata("kapferer2.Stochastic-Approximation.zeros")
kap2_sa_zero_edges <- load_network_rdata("kapferer2.Stochastic-Approximation.zeros_and_edges")
kap2_sa_mple <- load_network_rdata("kapferer2.Stochastic-Approximation.MPLE")
stats_kap2_sa <- load_stats_rdata("kapferer2.Stochastic-Approximation.zeros", "kapferer2.Stochastic-Approximation.zeros_and_edges", "kapferer2.Stochastic-Approximation.MPLE")
# equilibrium expectation
kap2_ee_zeros <- load_network_rdata("kapferer2.EE.zeros")
kap2_ee_zero_edges <- load_network_rdata("kapferer2.EE.zeros_and_edges")
kap2_ee_mple <- load_network_rdata("kapferer2.EE.MPLE")
stats_kap2_ee <- load_stats_rdata("kapferer2.EE.zeros", "kapferer2.EE.zeros_and_edges", "kapferer2.EE.MPLE")
```


# Analysis


## Estimated parameters

### ecoli base

```{r ecoli-estimated}

params_ecoli <- data.frame()
params_ecoli <- tabulate_parameters(params_ecoli, stats_ecoli_ee)
params_ecoli <- tabulate_parameters(params_ecoli, stats_ecoli_sa)
params_ecoli <- tabulate_parameters(params_ecoli, stats_ecoli_mcmle)
print(params_ecoli)
print.xtable(xtable(params_ecoli), file = "~/math5005/msresearch/outputs/subphase10/tables/ecoli1_params.txt")
```

### ecoli self

```{r}

self_params_ecoli <- data.frame()
self_params_ecoli <- tabulate_parameters(self_params_ecoli, self_stats_ecoli_ee)
self_params_ecoli <- tabulate_parameters(self_params_ecoli, self_stats_ecoli_sa)
self_params_ecoli <- tabulate_parameters(self_params_ecoli, self_stats_ecoli_mcmle)
print(self_params_ecoli, row.names = F)
print.xtable(xtable(self_params_ecoli), file = "~/math5005/msresearch/outputs/subphase10/tables/self_ecoli_params.txt")
```

### Kap 1

```{r}
params_kap1 <- data.frame()
params_kap1 <- tabulate_parameters(params_kap1, stats_kap1_ee)
params_kap1 <- tabulate_parameters(params_kap1, stats_kap1_sa)
# params_kap1 <- tabulate_parameters(params_kap1, stats_kap1_mcmle)
print(params_kap1)
print.xtable(xtable(params_kap1), file = "~/math5005/msresearch/outputs/subphase10/tables/params_kap1.txt")
```


### Kap 2
```{r}
params_kap2 <- data.frame()
params_kap2 <- tabulate_parameters(params_kap2, stats_kap2_ee)
params_kap2 <- tabulate_parameters(params_kap2, stats_kap2_sa)
# params_kap1 <- tabulate_parameters(params_kap1, stats_kap1_mcmle)
print.xtable(xtable(params_kap2), file = "~/math5005/msresearch/outputs/subphase10/tables/kap2_params.txt")
print(params_kap2)

```


## Runtimes

### E. Coli 1
```{r runtimes-ecoli}

df_ecoli <- data.frame()
df_ecoli <- tabulate_runtimes(df_ecoli, stats_ecoli_ee)
df_ecoli <- tabulate_runtimes(df_ecoli, stats_ecoli_sa)
df_ecoli <- tabulate_runtimes(df_ecoli, stats_ecoli_mcmle)
print(df_ecoli)
print.xtable(xtable(df_ecoli), file = "~/math5005/msresearch/outputs/subphase10/tables/ecoli1_runtimes.txt")
```

### E. Coli 2

```{r runtimes-ecoli-self}

df_self_ecoli <- data.frame()
df_self_ecoli <- tabulate_runtimes(df_self_ecoli, self_stats_ecoli_ee)
df_self_ecoli <- tabulate_runtimes(df_self_ecoli, self_stats_ecoli_sa)
df_self_ecoli <- tabulate_runtimes(df_self_ecoli, self_stats_ecoli_mcmle)
print(df_self_ecoli)
print.xtable(xtable(df_self_ecoli), file = "~/math5005/msresearch/outputs/subphase10/tables/self_ecoli1_runtimes.txt")
```

### Kap 1

```{r}
df_kap1 <- data.frame()
df_kap1 <- tabulate_runtimes(df_kap1, stats_kap1_ee)
df_kap1 <- tabulate_runtimes(df_kap1, stats_kap1_sa)
# df_kap1 <- tabulate_runtimes(df_kap1, stats_kap1_mcmle)
print(df_kap1)
print.xtable(xtable(df_kap1), file = "~/math5005/msresearch/outputs/subphase10/tables/kap1_runtimes.txt")
```


### Kap 2

```{r}
df_kap2 <- data.frame()
df_kap2 <- tabulate_runtimes(df_kap2, stats_kap2_ee)
df_kap2 <- tabulate_runtimes(df_kap2, stats_kap2_sa)
# df_ecoli <- tabulate_runtimes(df_ecoli, stats_kap2_mcmle)
print(df_kap2)
print.xtable(xtable(df_kap2), file = "~/math5005/msresearch/outputs/subphase10/tables/kap2_runtimes.txt")
```

## Traceplots

### Ecoli data set

#### EE

theta0 = zeros
```{r}
mcmc.diagnostics(ecoli_ee_zeros)
```

theta0 = zeros and edges
```{r}
mcmc.diagnostics(ecoli_ee_zero_edges)
```

theta0 = mple

```{r ecoli1-ee}
mcmc.diagnostics(ecoli_ee_mple)
```
#### SA

theta0 = zeros
```{r}
mcmc.diagnostics(ecoli_sa_zeros)
```

theta0 = zeros and edges
```{r}
mcmc.diagnostics(ecoli_sa_zero_edges)
```

theta0 = mple

```{r ecoli1-sa}
mcmc.diagnostics(ecoli_sa_mple)
```

#### MCMLE

theta0 = zeros
```{r}
mcmc.diagnostics(ecoli_mcmle_zeros)
```

theta0 = zeros and edges
```{r}
mcmc.diagnostics(ecoli_mcmle_zero_edges)
```

theta0 = mple

```{r ecoli1-mcmle}
mcmc.diagnostics(ecoli_mcmle_mple)
```


### Ecoli data set + self loop

#### EE 

theta0 = zeros
```{r}
mcmc.diagnostics(self_ecoli_ee_zeros)
```

theta0 = zeros and edges
```{r}
mcmc.diagnostics(self_ecoli_ee_zero_edges)
```

theta0 = mple

```{r ecoli2-ee}
mcmc.diagnostics(self_ecoli_ee_mple)
```

#### SA

theta0 = zeros
```{r}
mcmc.diagnostics(self_ecoli_sa_zeros)
```

theta0 = zeros and edges
```{r}
mcmc.diagnostics(self_ecoli_sa_zero_edges)
```

theta0 = mple
```{r ecoli2-sa}
mcmc.diagnostics(self_ecoli_sa_mple)
```

#### MCMLE

theta0 = zeros

```{r}
mcmc.diagnostics(self_ecoli_mcmle_zeros)
```

theta0 = zeros and edges

```{r}
mcmc.diagnostics(self_ecoli_mcmle_zero_edges)
```

theta0 = mple

```{r ecoli2-mcmle}
mcmc.diagnostics(self_ecoli_mcmle_mple)
```
### Kapferer 1

#### EE

theta0 = 0
```{r}
mcmc.diagnostics(kap1_ee_zeros)
```
theta0 = edges
```{r}
mcmc.diagnostics(kap1_ee_zero_edges)
```
thete0 = mple
```{r}
mcmc.diagnostics(kap1_ee_mple)
```

#### SA

theta0 = 0
```{r}
mcmc.diagnostics(kap1_sa_zeros)
```
theta0 = edges
```{r}
mcmc.diagnostics(kap1_sa_zero_edges)
```
thete0 = mple
```{r}
mcmc.diagnostics(kap1_sa_mple)
```

#### Kapferer2 

#### EE

theta0 = 0
```{r}
mcmc.diagnostics(kap2_ee_zeros)
```
theta0 = edges
```{r}
mcmc.diagnostics(kap2_ee_zero_edges)
```
thete0 = mple
```{r}
mcmc.diagnostics(kap2_ee_mple)
```

#### SA

theta0 = 0
```{r}
mcmc.diagnostics(kap2_sa_zeros)
```
theta0 = edges
```{r}
mcmc.diagnostics(kap2_sa_zero_edges)
```
thete0 = mple
```{r}
mcmc.diagnostics(kap2_sa_mple)
```

